The abundance of Bullet-Groups in $\Lambda$CDM

### 2014-03-20 paper/ms.tex
removed velocity section
new darft with new figures
new numbers

We estimate the expected distribution of displacements between the two
dominant dark matter density peaks and between baryons and dark matter
in halos simulated within a full cosmological context. We use as a benchmark the
observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated 
baryonic component. We extend previous results by
studying two halo samples: groups and clusters. We find that $(55\pm 5)\%$
of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$133\pm21$\hkpc as observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $(82\pm 3)\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1$ to
$1.0\%$ of the groups should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for clusters this fraction rises to $(7\pm 3)\%$, consistent
with previous studies of dark matter to baryon separations. The
extension of the theoretical predictions and observational results
towards low mass bi-modal dark matter configurations opens up the
possibility for a new statistical test of  $\Lambda$CDM.

NOTA: tomar $186\pm30$ para el DM-DM displacement.

\end{abstract}

@@

In this Letter we make a study at four different redshifts $z=0.0,0.25,
0.5$ and $1.0$. First, we select all the host halos (i.e. halos that
are not inside a larger halo) with circular velocities $V_{\rm c}\geq
300$\kms ($\geq 1\times 10^{13}\\hMsun$). Then, we select the sub-halos with
circular velocities $V_{\rm c}\geq 75$\kms ($\geq 5\times
10^{10}\hMsun$). The objective is to use this sub-halos as the tracer
of the sub-dominant dark matter clump in the merging cluster, i.e. the
bullets.

For the redshifts $z=0.0,0.25, 0.5$ and $1.0$ we find $10041, 10346,
10554$ and $10382$ host halos and $157853, 177331, 195072, 225188$
sub-halos, respectively. These two sets (host halos and sub-halos)
constitute the basis for our analysis. We associate each host halo
to its most massive sub-halo, each one of these pairs (host \& sub-halo)
is considered as a potential Bullet system and is kept for the
analysis described in the next Section.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_1.eps}
\end{center}
\caption{
  Integrated probability distribution for the displacement between the
  center of the host halo and its dominant sub-halo. The left panel
  shows the results in terms of the physical displacements while the
  right panel shows the displacements normalized by the virial radius
  of the host halo. The continuous (dashed) line corresponds to the halos in the
  cluster (group).
  The vertical lines show the mean value and uncertainties ($133\pm
  21$\hkpc) in the  separation between the two dark matter clumps
  estimated in \citet{Gastaldello} for the \bullg. Between $50\%$
  to $60\%$ of the groups show a displacement equal or larger than
  this observational benchmark.}
\label{fig:displacement}
\end{figure*}

@@

The panel with the projected 2D physical displacements also shows a
vertical stripe with the estimated displacement for the Bullet-group reported by
\cite{Gastaldello}. In the group sample we see that a fraction of
$50\%$ to $60\%$  should present a displacement equal than the
estimate for \bullg; in the cluster sample this fraction increases to
$80\%$-$85\%$. The uncertainties in this estimates are derived from
the uncertainties in the displacement measurement for \bullg.This
fraction is naturally higher in more massive systems because they are
larger in size. Normalizing the displacements by the virial radius,
right panel Figure \ref{fig:displacement}, we see that the
distribution is the same regardless of the sample and the redshift.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_4.eps}
\end{center}
\caption{2D histograms in the plane $\mu$-$D_{\rm off}$. The left
  panel corresponds to clusters and the right panel to groups. The data
  used to construct this histogram includes the halos at all redshifts.}
\label{fig:geometry}
\end{figure*}

%In Figure \ref{fig:velocities} we present the integrated probability
%of the relative peculiar velocities of the sub-halos with respect to
%the host halo. This velocity is presented in the left panel in
%physical units and in the right panel as a fraction of the circular
%velocity of the host halo.

%The panel with the normalized velocities shows that the distribution
%of sub-halo velocities is close to universal. Regardless of the mass
%of the host halo and the redshift the integrated distributions lie
%very close to each other.  The universality of this profile also holds
%for the group mass range. extending the result found by
%\citep{Hayashi2006} for cluster mass halos.

%The median of this distribution is located at $v/V_{\rm
%  c,host}=1.1$. We also note the strong break at $v/V_{\rm
%  c,host}=3.0$ present in the data from the group sample, which
%allows us to probe fractions on the order of $10^{-4}$.  This break is
%located close the escape velocity of $v/V_{\rm c,host}$ for dark
%matter halos following a NFW profile with a concentration value
%$c\approx 6$ \citep{Hayashi2006}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{figure_5.eps}
\end{center}
\caption{Integrated probability distribution for the estimated
  baryonic displacements in the group and cluster samples. Continous (dashed)
  lines correspond to clusters (groups). The continuous black line
  marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and baryons. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.} 
\label{fig:baryonic_displacements}
\end{figure}

@@

Recently \citep{Foex2013} presented an analysis of $80$ galaxy groups
from the SL2S sample. From the light distribution, only $34$ objects
($\sim 42\%$) have regular isophotes, $33$ had ellongated
isophotes (hints of merging system), and $13$ ($\sim 16\%$) had a
clear bi-modal light distribution.

The bi-modal objects are defined to have at least a clear second
luminosity peak within $350$\hkpc from the main halo, as traced by the
strong lensing system.  The lowest separation in those systems is
$64$ \hkpc and the average is $145\pm 52$\hkpc. Comparing this average
displacement against the integratied distribution for the group sample (left
panel, Figure \ref{fig:displacement}) one would expects between $40\%$
to $80\%$ of DM halos to present those displacements. This fraction is
certainly larger than the $\approx 16\%$ inferred from the
observational sample.

We have taken the results from the simulation and selected all the
objects in the group sample with a DM-DM displacement less than
$350$\hkpc but larger than $64$\hkpc to find XXX objects in
total. From this subsample we find that XXX have a DM-baryon
displacement larger than $87$\hkpc.

This rough comparison hints that it is possible to find a large fraction
of multi-modal groups with a DM-baryon displacement in a $\Lambda$CDM
cosmology. A proper comparison must take into account all the
observational uncertainties and biases to derive a stronger prediction from the
simulation, something that is beyond the scope of this Letter.

@@

In this Letter we estimate fraction of galaxy groups and clusters that
can present observational features associated to a bullet-like
event. This is motivated by the recent observational results of
\citep{Gastaldello} where a system (\bullg) on the mass range $1\times
10^{14}$\hMsun\ and velocity dispersion $650$\kms was reported to
feature a displacement between its baryonic (hot gas) and dark matter
components.

@@

Our main results is that that a fraction of $50\%$-$60\%$ of the halos
in the group sample present displacement equal or larger than the
observed displacement for \bullg. For halos in the cluster sample this
fraction increases to $80\%$-$85\%$. We also derive an
estimate for the displacement between the DM and the baryonic
component. In the group sample $0.1\%$-$1.0\%$ of the halos show a
displacement equal or larger than  the  measurements of \bullg\ by
\citep{Gastaldello}. In the cluster sample this fraction rises to
$4\%$-$10\%$.

We also find distributions for the DM separation and the velocity of
the bullet through its host. If these quantities are normalized by the
virial radius and the circular velocity, respectively, we arrive at
distributions close to universal that are similar for the two halo
samples at all redshifts.

For the case of \bullg\ a fair comparison is achieved against cluster
sample, which has statistics dominated by objects of similar mass. In
this case we conclude that the existence of such configuration is
highly probable in $\Lambda$CDM (4$\%$ to $10\%$ abundance). In turn,
for the same separation the fraction for groups is lower (0.1$\%$ to
$1\%$). Taking into account that the difference in spatial abundance
between these two samples is on the order of a factor of $10$, one can
conclude that the absolute number of groups and clusters presenting a
DM-baryon displacement larger or equal to \bullg\ should be of the
same order.

An observational possibility opens up with surveys such as SL2S that
targetted a large number of groups and estimate its multi-modal nature
from lensing analysis \citep{Foex2013}, an approach that can be
exploited with upcoming lensing surveys (e.g. with the {\it Euclid}
satellite).

>>>>

We estimate the expected distribution of displacements between the two
dominant dark matter density peaks and between baryons and dark matter
in halos simulated within a full cosmological context. We use as a benchmark the
observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated 
baryonic component. We extend previous computational results by
studying two halo samples: groups and clusters. We find that $(50 \pm 10)\%$
of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$186 \pm 30$\hkpc as observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $(70 \pm 10)\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1$ to
$1.0\%$ of the groups should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for clusters this fraction rises to $(7\pm 3)\%$, consistent
with previous studies of dark matter to baryon separations. In terms
of absolute numbers densities we find that the expected value for
bullet groups is tree times larger than the expected value for bullet
groups. The extension of the theoretical predictions and observational
results towards low mass bi-modal dark matter configurations opens up
the possibility for a new statistical test of  $\Lambda$CDM.  
\end{abstract}

@@

In this Letter we make a study at four different redshifts $z=0.0,0.25,
0.5$ and $1.0$. First, we select all the host halos (i.e. halos that
are not inside a larger halo) with circular velocities $V_{\rm c}\geq
300$\kms ($\geq 1\times 10^{13} \hMsun$). Then, we select the sub-halos with
circular velocities $V_{\rm c}\geq 75$\kms ($\geq 5\times
10^{10}\hMsun$). The objective is to use this sub-halos as the tracer
of the sub-dominant dark matter clump in the merging cluster, i.e. the
bullets.

These two sets (host halos and sub-halos)
constitute the basis for our analysis. We associate each host halo
to its most massive sub-halo, each one of these pairs (host \& sub-halo)
is considered as a potential Bullet system and is kept for the
analysis described in the next Section. The number of systems at each
redshift is listed on the first arrow of Table \ref{table:numbers}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_1.eps}
\end{center}
\caption{
  Integrated probability distribution for the displacement between the
  center of the host halo and its dominant sub-halo. The left panel
  shows the results in terms of the physical displacements while the
  right panel shows the displacements normalized by the virial radius
  of the host halo. The continuous (dashed) line corresponds to the halos in the
  cluster (group).
  The vertical lines show the mean value and uncertainties ($186\pm
  30$\hkpc) in the  separation between the two dark matter clumps
  estimated in \citet{Gastaldello} for the \bullg. Between $40\%$
  to $60\%$ of the groups show a displacement equal or larger than
  this observational benchmark. This fraction rises to $60\%$ and
  $80\%$ in clusters.}
\label{fig:displacement}
\end{figure*}

@@

The panel with the projected 2D physical displacements also shows a
vertical stripe with the estimated displacement for the Bullet-group reported by
\cite{Gastaldello}. In the group sample we see that a fraction of
$40\%$ to $60\%$  should present a displacement equal than the
estimate for \bullg; in the cluster sample this fraction increases to
$70\%$-$80\%$. The uncertainties in this estimates are derived from
the uncertainties in the displacement measurement for \bullg.This
fraction is naturally higher in more massive systems because they are
larger in size. Normalizing the displacements by the virial radius,
right panel Figure \ref{fig:displacement}, we see that the
distribution is the same regardless of the sample and the redshift.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_3.eps}
\end{center}
\caption{2D histograms in the plane $\mu$-$D_{\rm off}$. The left
  panel corresponds to clusters and the right panel to groups. The data
  used to construct this histogram includes the halos at all redshifts.}
\label{fig:geometry}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{figure_4.eps}
\end{center}
\caption{Integrated probability distribution for the estimated
  baryonic displacements in the group and cluster samples. Continous (dashed)
  lines correspond to clusters (groups). The continuous black line
  marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and baryons. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.} 
\label{fig:baryonic_displacements}
\end{figure}

@@

Recently \citep{Foex2013} presented an analysis of $80$ galaxy groups
in the SL2S sample. From the light distribution, only $34$ objects
($\sim 42\%$) have regular isophotes, $33$ had ellongated
isophotes (hints of merging system), and $13$ ($\sim 16\%$) had a
clear bi-modal light distribution.  The system \bullg\ is one of these
$13$ systems.

The bi-modal objects are defined to have at least a clear second
luminosity peak within $350$\hkpc from the main halo, as traced by the
strong lensing system.  The lowest separation in those systems is
$64$ \hkpc and the average is $145\pm 52$\hkpc.

We now make a comparison of these fractions against the results of our
simulations. The results are summarized in Table
\ref{table:numbers}. The first row indicate the total number of halos
in each sample, at each redshift. The second row shows the number of
objects with DM-DM displacement $46\hkpc< d_{\rm 2D}< 350$\hkpc. The
last row indicates the number of objects in the previous sub-sample
with a  $d_{\rm 2D}^{\rm bar}>87$\hkpc.

Considering that the statistics for the cluster sample are dominated
by objects in the mass range of the \bullg we make a comparison
against this sample. The table shows that $\sim 40\%$ of the 
clusters are expected to have large DM-DM displacements, this is a
factor of $\sim 2$ larger than the observational estimate by
\citep{Foex2013}. However, roughly $\sim 7\%$ of these systems present a
displacement equal or larger than the observed in the \bullg, a
fraction that is compatible with the $1/13\sim 0.07$ fraction in the
observational sample.

This rough comparison hints that it is possible to find a large fraction
of multi-modal groups with a DM-baryon displacement in a $\Lambda$CDM
cosmology and that our estimates for the relative number of bullet
systems with respect to multi-modal systems is compatible with
observations.

A proper comparison must take into account all the observational
uncertainties, biases, mixes between our two populations (groups \&
clusters at different redshifts) to derive a stronger prediction from
the simulation, something that is beyond the scope of this Letter.

\begin{table*}
\begin{center}
\begin{tabular}{ccccccccc}\hline\hline
        & Groups & Groups & Groups & Groups & Clusters & Clusters & Clusters & Clusters\\
  & $z=0$   & $z=0.25$   & $z=0.5$   & $z=1.0$   & $z=0$   & $z=0.25$   & $z=0.5$   & $z=1.0$ \\
  & ($\# $)    & ($\# $)    & ($\# $)    & ($\# $)    & ($\# $)    & ($\# $)    & ($\# $)    & ($\# $)  \\\hline
Full Sample & 9641 & 9984 & 10244 & 10190 & 400  & 363 & 310 & 192 \\
$64 <d_{\rm 2D}/\hkpc< 350$ & 6188 & 6422 & 6635 & 6933 & 151 &
141 & 120 & 99\\
($64 <d_{\rm 2D}/\hkpc< 350 $) \& ($d_{\rm 2D}^{\rm bar}/\hkpc >87$) & 14 & 25 & 44 & 35 & 8 & 9 & 13 & 8 \\\hline\hline
\end{tabular}
\caption{Absolute number of objects in the groups and clusters sample
  at all redshifts for different selection criteria. }
\label{table:numbers}
\end{center}
\end{table*}

@@

In this Letter we estimate the fraction of galaxy groups and clusters that
can present observational features associated to a bullet-like
event. This is motivated by the recent observational results of
\citep{Gastaldello} where a system (\bullg) on the mass range $1\times
10^{14}$\hMsun\ and velocity dispersion $650$\kms was reported to
feature a displacement between its baryonic (hot gas) and dark matter
components.

@@

Our main results is that that a fraction of $40\%$-$60\%$ of the halos
in the group sample present displacement equal or larger than the
observed displacement for \bullg. For halos in the cluster sample this
fraction increases to $60\%$-$80\%$. We also derive an
estimate for the displacement between the DM and the baryonic
component. In the group sample $0.1\%$-$1.0\%$ of the halos show a
displacement equal or larger than  the  measurements of \bullg\ by
\citep{Gastaldello}. In the cluster sample this fraction rises to
$4\%$-$10\%$.

For the case of \bullg\ a fair comparison is achieved against cluster
sample, which has statistics dominated by objects of similar mass. In
a rough comparison using the selection criteria by \citep{Gael2013,
  Gastaldello} we find that the relative number of bullet groups and
systems with large DM-DM displacements is consistent with
observations; both are in the range $\sim 7\%$. In terms of absolute
numer densities, we expect $\sim 6.0\times 10^{-7}$ Mpc$^{-3}$ Bullet
groups similar to \bullg; three times the expected value for Bullet
clusters.

We conclude that a new observational possibility opens up with surveys
such as SL2S that targett a large number of groups and estimate its
multi-modal nature from lensing analysis \citep{Foex2013}, an approach
that can be exploited with upcoming lensing surveys (e.g. with the
{\it Euclid} satellite).
